Gusto: a toolkit for compatible finite element dynamical cores

### 2025-08-06 source/paper.tex
small changes to compile and fix referencing

In line with the theory outlined in the introduction, we always choose
the velocity $\MM{u} \in \ring{\mathbb{V}}^{HDiv}$ with
$\ring{\mathbb{V}}^{HDiv} \subset \mathbb{V}^{HDiv} \subset
H(\text{div})$ and, as mentioned above, the $\ring{}$ indicates that
the functions in $\ring{\mathbb{V}}^{HDiv}$ satisfy the no normal flow
boundary conditions. The field whose gradient appears in the
`pressure' gradient term is chosen to be in the space
$\mathbb{V}^{L^2}$ with $\mathbb{V}^{L^2} \subset L_2$, or, in the
case of the more complex nonlinear pressure gradient term for the
compressible Euler equations, it is a function of a variable in this
space. The compressible Boussinesq and compressible Euler equations
have a third prognostic variable, either buoyancy or potential
temperature, which is chosen to be in the space $\mathbb{V}_\theta$
which has degrees of freedom that are collocated with those of the
vertical velocity, mimicking the Charney-Phillips staggering as
discussed in **. These choices are governed by the properties required
of discretisation of the linearised equation sets (as discussed in
section ** and other refs). To solve the nonlinear equation sets we
need to introduce stable and accurate discretisations of the transport
terms and the nonlinear pressure gradient term in the compressible
Euler equation set. In addition, some test cases require viscous and
diffusive terms to generate converged solutions \ref{}. In the
following subsections we illustrate a non-exhaustive selection of the
options available in Gusto.

@@

\subsection{Terminator Toy chemistry}
The first results use the ``Terminator Toy'' test of \citet{lauritzen2015terminator}, which involves the transport of two coupled tracers $X$ and $X_2$ on the surface of a sphere.
The chemicals interact such that the sum $m_T=m_X+2m_{X_2}$ should be preserved, where $m_X$ and $m_{X_2}$ are the mixing ratios of the individual chemicals.
When positivity or monotonicity limiters are applied to the numerical transport scheme, this may break the correlation between the two species which will be highlighted by this test.
This test is also a demonstration of simple ``physics-dynamics coupling'', as the conversion term between the species can be handled separately from the transport terms. \\
\\
For this test, we use the DG upwind scheme described in Section \ref{sec:transport} to transport the species on an icosahedral sphere mesh, and apply the vertex-based limiter of \citet{kuzmin2010vertex}.
The term describing the chemical reaction is separated from the transport term using a simple sequential splitting, as described in Section \ref{sec:seq splitting}. \\
\\
Figure \ref{fig:terminator} shows the mixing ratios of the two individual species at the initial time and following a single rotation around the sphere by a deformational flow.
The change in the sum of the species is also displayed, with differences from zero remaining close to machine precision, which demonstrates that this transport scheme does preserve the correlation between the tracers.

@@

\begin{sidewaystable*}[tbp!]
\caption{Details of the configurations used for the test cases in Section \ref{sec: results}.}
\begin{tabular}{l|ccccccc}
\tophline
Test Case & Equation Set & Timestepper & Finite Elements & \begin{tabular}{@{}c@{}}Horizontal \\ Mesh\end{tabular} & \begin{tabular}{@{}c@{}}Number \\ of  \ Layers\end{tabular} & Time step & Final Time \\
\middlehline
Terminator Toy & \begin{tabular}{@{}c@{}}Tracer transport \\ with chemistry\end{tabular} & Sequential Splitting & BDM1 & I16 & None & 450 s & 12 days \\
Steady Geostrophic Flow & Linear Shallow Water & Multiple & BDM1 & I32 & None & ?? & 1 day \\
Steady Geostrophic Flow & Linear Shallow Water & Trapezium Rule & Multiple & ?? & None & ?? & 1 day \\
Unstable Jet & Shallow Water & SIQN & BDM1 and RT1 & I64 and C100 & None & 150 s & 6 days \\
Density Current & Compressible Euler & SIQN & RT0 and RT1 & 100 m and 200 m & 64 and 32 & 1 s & 900 s \\
Rising Thermal & Moist Compressible Euler & SIQN & RT1 & $\Delta x = \Delta y = \frac{10}{81}$ km & 81 & 2 s & 1000 s \\
Baroclinic Wave & Compressible Euler & SIQN & RT1 & C32 & 15 & 900 s & 10 days
\bottomhline
\end{tabular}
\belowtable{Notation: RT$x$ refers to the de Rham complex that uses the Raviart-Thomas elements on quadrilateral cells for the HDiv function space, and results in an $x$-th order polynomials for the $L_2$ space. Similarly, BDM$x$ refers to the Brezzi-Douglas-Marini de Rham complex on triangular cells.
SIQN denotes the Semi-Implicit Quasi-Newton timestepper described in Section \ref{sec:siqn}. When describing horizontal meshes of the sphere, I$y$ and C$y$ correspond to icosehedral-sphere and cubed-sphere meshes with $y$ cells per edge of the icosahedron or cube panel.}% Table Footnotes
\end{sidewaystable*}

>>>>

In line with the theory outlined in the introduction, we always choose
the velocity $\MM{u} \in \ring{\mathbb{V}}^{HDiv}$ with
$\ring{\mathbb{V}}^{HDiv} \subset \mathbb{V}^{HDiv} \subset
H(\text{div})$ and, as mentioned above, the $\ring{}$ indicates that
the functions in $\ring{\mathbb{V}}^{HDiv}$ satisfy the no normal flow
boundary conditions. The field whose gradient appears in the
`pressure' gradient term is chosen to be in the space
$\mathbb{V}^{L^2}$ with $\mathbb{V}^{L^2} \subset L_2$, or, in the
case of the more complex nonlinear pressure gradient term for the
compressible Euler equations, it is a function of a variable in this
space. The compressible Boussinesq and compressible Euler equations
have a third prognostic variable, either buoyancy or potential
temperature, which is chosen to be in the space $\mathbb{V}_\theta$
which has degrees of freedom that are collocated with those of the
vertical velocity, mimicking the Charney-Phillips staggering as
discussed in **. These choices are governed by the properties required
of discretisation of the linearised equation sets (as discussed in
section ** and other refs). To solve the nonlinear equation sets we
need to introduce stable and accurate discretisations of the transport
terms and the nonlinear pressure gradient term in the compressible
Euler equation set. In addition, some test cases require viscous and
diffusive terms to generate converged solutions (for example
\citet{straka1993numerical}). In the following subsections we
illustrate a non-exhaustive selection of the options available in
Gusto.

@@

\subsection{Terminator Toy chemistry}
The first results use the ``Terminator Toy'' test of \citet{lauritzen2015terminator}, which involves the transport of two coupled tracers $X$ and $X_2$ on the surface of a sphere.
The chemicals interact such that the sum $m_T=m_X+2m_{X_2}$ should be preserved, where $m_X$ and $m_{X_2}$ are the mixing ratios of the individual chemicals.
When positivity or monotonicity limiters are applied to the numerical transport scheme, this may break the correlation between the two species which will be highlighted by this test.
This test is also a demonstration of simple ``physics-dynamics coupling'', as the conversion term between the species can be handled separately from the transport terms. \\
\\
For this test, we use the DG upwind scheme described in Section \ref{subsec: DG transport} to transport the species on an icosahedral sphere mesh, and apply the vertex-based limiter of \citet{kuzmin2010vertex}.
The term describing the chemical reaction is separated from the transport term using a simple sequential splitting, as described in Section \ref{sec:seq splitting}. \\
\\
Figure \ref{fig:terminator} shows the mixing ratios of the two individual species at the initial time and following a single rotation around the sphere by a deformational flow.
The change in the sum of the species is also displayed, with differences from zero remaining close to machine precision, which demonstrates that this transport scheme does preserve the correlation between the tracers.

@@

\begin{sidewaystable*}[tbp!]
\caption{Details of the configurations used for the test cases in Section \ref{sec: results}.}
\begin{tabular}{l|ccccccc}
\tophline
Test Case & Equation Set & Timestepper & Finite Elements & \begin{tabular}{@{}c@{}}Horizontal \\ Mesh\end{tabular} & \begin{tabular}{@{}c@{}}Number \\ of  \ Layers\end{tabular} & Time step & Final Time \\
\middlehline
Terminator Toy & \begin{tabular}{@{}c@{}}Tracer transport \\ with chemistry\end{tabular} & Sequential Splitting & BDM1 & I16 & None & 450 s & 12 days \\
Steady Geostrophic Flow & Linear Shallow Water & Multiple & BDM1 & I32 & None & ?? & 1 day \\
Steady Geostrophic Flow & Linear Shallow Water & Trapezium Rule & Multiple & ?? & None & ?? & 1 day \\
Unstable Jet & Shallow Water & SIQN & BDM1 and RT1 & I64 and C100 & None & 150 s & 6 days \\
Density Current & Compressible Euler & SIQN & RT0 and RT1 & 100 m and 200 m & 64 and 32 & 1 s & 900 s \\
Rising Thermal & Moist Compressible Euler & SIQN & RT1 & $\Delta x = \Delta y = \frac{10}{81}$ km & 81 & 2 s & 1000 s \\
Baroclinic Wave & Compressible Euler & SIQN & RT1 & C32 & 15 & 900 s & 10 days \\
\bottomhline
\end{tabular}
\belowtable{Notation: RT$x$ refers to the de Rham complex that uses the Raviart-Thomas elements on quadrilateral cells for the HDiv function space, and results in an $x$-th order polynomials for the $L_2$ space. Similarly, BDM$x$ refers to the Brezzi-Douglas-Marini de Rham complex on triangular cells.
SIQN denotes the Semi-Implicit Quasi-Newton timestepper described in Section \ref{sec:siqn}. When describing horizontal meshes of the sphere, I$y$ and C$y$ correspond to icosehedral-sphere and cubed-sphere meshes with $y$ cells per edge of the icosahedron or cube panel.}% Table Footnotes
\end{sidewaystable*}
